---
title: "Machine learning paradigms and workflows -- Part 01"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "Overview of ML paradigms and workflows in R"
draft: true
---

# Pre-lecture activities

::: callout-tip

In advance of class, please install

* `tidymodels` - collection of packages for modeling and machine learning using tidyverse principles

You can do this by calling

```{r}
#| eval: false
install.packages("tidymodels")
```

And load the package using: 

```{r}
#| message: false
library(tidymodels)
```

In addition, please 

- Check out the [Getting Started](https://www.tidymodels.org/start/) guide in `tidymodels`

:::


# Lecture 

## Acknowledgements

Material for this lecture was borrowed and adopted from

- [An Introduction to Statistical Learning](https://www.statlearning.com)

## Learning objectives

::: callout-note
### Learning objectives

**At the end of this lesson you will:**

-  State the differences between machine learning paradigms including supervised, unsupervised, semi-supervised, and reinforcement Learning
-  Be familiar with logistic regression and random forests as two types of classification methods in supervised learning  
-  Recognize evaluation techniques
-  Use `tidymodels` as collection of packages for modeling and machine learning using `tidyverse` principles 

:::



## Slides 

- [Lecture 07: Machine learning paradgims -- Part 01]()

# Class activity 

For the rest of the time in class, you and your team will work on the final project. Stephanie will walk around to answer questions and happy to help in any way! 


# Post-lecture

## Additional practice

Here are some additional practice questions to help you think about the material discussed.

### Question 1

Using `penguins` dataset discussed in class, we will practice more model 
building by experimenting with other features. In our first logistic 
regression, we only used two features (`bill_length_mm` and 
`flipper_length_mm`). In the second, we added `body_mass_g` and `species`.

Now, fit a third logistic regression model that uses only `bill_length_mm` 
and `body_mass_g` as input features (X). Compute the accuracy and Kappa metrics 
using `metrics()`. Compare the results to both of the previous logistic 
regression models. Which combination of predictors yields the best performance, 
and why might that be? The purpose of this question is to help build intuition 
about how including or excluding features affects predictive accuracy and 
model interpretability.

::: {.callout-tip collapse="true"}
### Solution

```{r}
library(tidymodels)
library(palmerpenguins)

penguins <- na.omit(penguins)

# Logistic regression using only bill_length_mm and body_mass_g
logistic_model_q1 <- 
  logistic_reg() %>% 
  set_engine("glm") %>% 
  fit(sex ~ bill_length_mm + body_mass_g, data = penguins)

# Examine model coefficients
tidy(logistic_model_q1)

# Make predictions
logistic_preds_q1 <- predict(logistic_model_q1, penguins) %>%
  bind_cols(penguins)

# Evaluate model performance
logistic_preds_q1 %>%
  metrics(truth = sex, estimate = .pred_class)
```

We saw that `flipper_length_mm` was not an informative feature, but it seems 
like `body_mass_g` is an informative variable! When we drop uninformative 
features and add informative features to the model, the predictive accuracy 
improves. It also helps model interpretability because we are keeping only a
minmial set of features in the model.

:::

### Question 2

In class, we compared logistic regression to random rorest models. We also 
computed the accuracy and Kappa for both the random forest and logistic 
regression models. Which performed better on this dataset? Why might the 
random forest outperform (or underperform) logistic regression here?

::: {.callout-tip}

Use `vip::vip(rf_fit)` to visualize which features the random forest 
considered most important. If you do this, you will need to modify 
`set_engine("ranger", importance = "impurity")` to calculate `importance`

:::

The purpose of this question is to compare model types and interpret 
variable importance in tree-based models.

::: {.callout-tip collapse="true"}
### Solution

```{r}
library(vip)

# Logistic regression with multiple predictors
logistic_model_q2 <- 
  logistic_reg() %>%
  set_engine("glm") %>%
  fit(sex ~ bill_length_mm + flipper_length_mm + body_mass_g + species, 
      data = penguins)

# Random forest with same predictors
rf_spec_q2 <- rand_forest(mtry = 3) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_fit_q2 <- rf_spec_q2 %>%
  fit(sex ~ bill_length_mm + flipper_length_mm + body_mass_g + species, 
      data = penguins)

# Evaluate both models
logistic_preds_q2 <- predict(logistic_model_q2, penguins) %>%
  bind_cols(penguins)

rf_preds_q2 <- predict(rf_fit_q2, penguins) %>%
  bind_cols(penguins)

metrics(logistic_preds_q2, truth = sex, estimate = .pred_class)
metrics(rf_preds_q2, truth = sex, estimate = .pred_class)

# Visualize feature importance for random forest
vip(rf_fit_q2)
```

We see that random forest performed better than logistic regression. 
This is because random forest can model complex, non-linear decision 
boundaries, while logistic regression assumes a linear relationship 
between the features and the log-odds of the outcome.

Also random forest can help with overfitting. It creates an ensemble of 
multiple decision trees and averaging their predictions. Therefore, random 
forest is more robust and less prone to overfitting than a single decision tree.

It also provides a measure of which features are most important for making 
predictions, which can be useful for understanding the data.

:::

### Question 3

The random forest model was trained and evaluated on the entire dataset. 
That’s not ideal for assessing generalization of how well the model will 
perform on data it was not trained on. 

Next, use `initial_split()` from `rsample` to create a training (80%) 
and testing (20%) dataset. Fit your random forest on the training data and 
compute metrics on the test data. Compare the test accuracy to the earlier 
“full dataset” accuracy. What does the difference tell you about 
potential overfitting? The purpose of this question is to practice a strategy 
that can improve the model-building workflow that we will learn about 
in our next lecture. 

::: {.callout-tip collapse="true"}
### Solution

```{r}
set.seed(123)
peng_split <- initial_split(penguins, prop = 0.8)
peng_train <- training(peng_split)
peng_test  <- testing(peng_split)

# Reuse the same random forest specification
rf_spec_q3 <- rand_forest(mtry = 3) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Fit model on training data
rf_fit_q3 <- rf_spec_q3 %>%
  fit(sex ~ bill_length_mm + flipper_length_mm + body_mass_g + species, 
      data = peng_train)

# Evaluate on test data
rf_preds_q3 <- predict(rf_fit_q3, peng_test) %>%
  bind_cols(peng_test)

metrics_train <- predict(rf_fit_q3, peng_train) %>%
  bind_cols(peng_train) %>%
  metrics(truth = sex, estimate = .pred_class)

metrics_test <- rf_preds_q3 %>%
  metrics(truth = sex, estimate = .pred_class)

metrics_train
metrics_test
```

The performance dropped in the test data, therefore we are potentially
overfitting when we fit our models. 

:::