---
title: "Machine learning paradigms and workflows -- Part 02"
author: 
  - name: Stephanie Hicks
    url: https://stephaniehicks.com
    affiliation: Department of Biostatistics, Johns Hopkins
    affiliation_url: https://publichealth.jhu.edu
description: "Overview of ML paradigms and workflows in R"
draft: true
---

# Pre-lecture activities

::: {.callout-tip}

We will continue with [`tidymodels`](https://www.tidymodels.org/start/), 
so there is nothing else to install for the lecture. However, if you 
want to see a different linear regression model workflow with the `penguins` 
dataset, checkout this tutorial: 

- [Regression in tidymodels](https://cmu-delphi.github.io/delphi-tooling-book/tidymodels-regression.html)

:::


# Lecture 

## Acknowledgements

Material for this lecture was borrowed and adopted from

- [An Introduction to Statistical Learning](https://www.statlearning.com)

## Learning objectives

::: callout-note
### Learning objectives

**At the end of this lesson you will:**

-  Use `tidymodels` as collection of packages for modeling and machine learning using `tidyverse` principles
-  Be able to fit a linear regression model and a random forest model to be used for prediction in supervised learning 
-  Be able to split data into "train" and "test" to reduce overfitting and improve model performance

:::

## Slides 

- [Lecture 08: Machine learning paradgims -- Part 02]()

# Class activity 

For the rest of the time in class, we will practice implementing more ML
models with the same `penguins` dataset. 

::: callout-note
### Objectives of the activity

- Extend the linear regression example using additional predictors and compare performance
- Understand how a playing with hyperparameters can change model performance 
- Explore whether adding interaction terms improves predictions

:::

For this in-class activity, you can work alone or find a partner. I encourage you 
to find a partner so you can discuss the questions asked below. 

## Part 1

Using `penguins` dataset discussed in class, we will first extend the linear 
regression example using additional predictors and compare the model
performances. Using this dataset, build two linear regression models predicting
`body_mass_g` using:

- Model A: `bill_length_mm` + `species`
- Model B: `bill_length_mm` + `flipper_length_mm` + `species` + `sex`

Use the same train/test split for both models. Then, compare RMSE and R^2 
between Model A and Model B on:

- the training set
- the test set

Think about and discuss these questions with a partner (or to yourself): 

- Which model performs better?
- Did adding predictors help?
- Is there evidence of overfitting?

::: {.callout-tip collapse="true"}
### Solution

First, we will load libraries and split the data into train and test
```{r}
#| message: false
# shared setup
library(tidymodels)
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(vip)

# feature engineering
penguins <- na.omit(penguins)

# create a single train/test split used for all problems
set.seed(123)
p_split <- initial_split(penguins, prop = 0.8)
p_train <- training(p_split)
p_test  <- testing(p_split)

# metrics set we will use later on
reg_metrics <- metric_set(rmse, rsq, mae)
```

Next, we will compare Model A vs Model B 

```{r}
# Model A: bill_length_mm + species
model_a <- linear_reg() %>% set_engine("lm") %>% set_mode("regression") %>%
  fit(body_mass_g ~ bill_length_mm + species, data = p_train)

# Model B: bill_length_mm + flipper_length_mm + species + sex
model_b <- linear_reg() %>% set_engine("lm") %>% set_mode("regression") %>%
  fit(body_mass_g ~ bill_length_mm + flipper_length_mm + species + sex, data = p_train)

# predictions and metrics on training set
pred_a_train <- predict(model_a, p_train) %>% bind_cols(p_train)
pred_b_train <- predict(model_b, p_train) %>% bind_cols(p_train)

metrics_a_train <- pred_a_train %>% reg_metrics(truth = body_mass_g, estimate = .pred)
metrics_b_train <- pred_b_train %>% reg_metrics(truth = body_mass_g, estimate = .pred)

# predictions and metrics on test set
pred_a_test <- predict(model_a, p_test) %>% bind_cols(p_test)
pred_b_test <- predict(model_b, p_test) %>% bind_cols(p_test)

metrics_a_test <- pred_a_test %>% reg_metrics(truth = body_mass_g, estimate = .pred)
metrics_b_test <- pred_b_test %>% reg_metrics(truth = body_mass_g, estimate = .pred)

# print results
cat("=== Model A: Training metrics ===\n"); print(metrics_a_train)
cat("\n=== Model B: Training metrics ===\n"); print(metrics_b_train)
cat("\n=== Model A: Test metrics ===\n"); print(metrics_a_test)
cat("\n=== Model B: Test metrics ===\n"); print(metrics_b_test)

# compare observed vs predicted on test set for Model B (more features)
ggplot(pred_b_test, aes(x = body_mass_g, y = .pred, color = species)) +
  geom_point(alpha = 0.8) +
  geom_abline(linetype = "dashed") +
  labs(title = "Model B: Observed vs Predicted on Test Set",
       x = "Observed body_mass_g", y = "Predicted .pred")
```

Generally: 

- if Model B improves test rmse (and not only train rmse), added predictors help generalization
- if Model B reduces train rmse a lot, but test rmse does not improve (or worsens), that suggests overfitting

:::

## Part 2 

In this question, we will use the training set from above, but now we will 
switch to random forests. There is an argument called `mtry = 3` used to identify 
the number of features sampled for each tree. Let's try changing it and we can 
"tune" the this parameter (also known as a "hyperparameter"). We will ask if this
changes the model's performance. 

Using the training dataset, fit three random forest models with 
different `mtry` values:

- `mtry = 2`
- `mtry = 3` 
- `mtry = 5`

For each model:

- Fit on the training data
- Evaluate on the test data (e.g. rmse and rsq)
- Plot `mtry` vs test rmse.

Think about and discuss these questions with a partner (or to yourself): 

- Which `mtry` yields the lowest test rmse?
- Why might increasing `mtry` help or hurt performance?


::: {.callout-tip collapse="true"}
### Solution

```{r}
library(purrr)
library(tidyr)

# model set up
rf_formula <- body_mass_g ~ bill_length_mm + flipper_length_mm + sex + species

# set up a vector of mtry values (will use purrr for this)
mtry_values <- c(2, 3, 5)

# function to fit random forest and return test metrics
fit_rf_and_eval <- function(mtry_val) {
  spec <- rand_forest(mtry = mtry_val, trees = 500) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("regression")
  
  fit_obj <- spec %>% fit(rf_formula, data = p_train)
  
  pred_test <- predict(fit_obj, p_test) %>% bind_cols(p_test)
  met_test <- pred_test %>% reg_metrics(truth = body_mass_g, estimate = .pred) %>%
    filter(.metric %in% c("rmse", "rsq", "mae")) %>%
    mutate(mtry = mtry_val, set = "test")
  
  # also return training metrics optionally
  pred_train <- predict(fit_obj, p_train) %>% bind_cols(p_train)
  met_train <- pred_train %>% reg_metrics(truth = body_mass_g, estimate = .pred) %>%
    filter(.metric %in% c("rmse", "rsq", "mae")) %>%
    mutate(mtry = mtry_val, set = "train")
  
  bind_rows(met_train, met_test)
}

results_list <- map(mtry_values, fit_rf_and_eval)
results_df <- bind_rows(results_list)

# make it into a nice table to compare all the metrics
print(results_df %>% pivot_wider(names_from = c(set, .metric), values_from = .estimate))
```

Finally, we plot `mtry` vs test rmse

```{r}
results_df %>%
  filter(set == "test", .metric == "rmse") %>%
  ggplot(aes(x = factor(mtry), y = .estimate, group = 1)) +
  geom_point(size = 3) + geom_line() +
  labs(title = "test rmse vs mtry (random forest)",
       x = "mtry", y = "test rmse")
```

Generally: 

- Small `mtry` reduces correlation between trees and may reduce overfitting
- Large `mtry` lets trees be more similar to full trees (may reduce bias but increase variance)

:::

## Part 3

In this question, fit two linear regression models to predict `body_mass_g`:

- Model without interaction: `body_mass_g ~ bill_length_mm + flipper_length_mm + species`
- Model with interaction: `body_mass_g ~ bill_length_mm * species + flipper_length_mm`

You want want to compare both models on the train and test rmse. Then, create a visualization of predicted vs actual body mass for the test set for each model.

Think about and discuss these questions with a partner (or to yourself): 

- Did adding interactions help?
- Why might interactions between `bill_length_mm` and `species` matter biologically?

::: {.callout-tip collapse="true"}
### Solution

```{r}
# model without interaction
model_no_int <- linear_reg() %>% set_engine("lm") %>% fit(
  body_mass_g ~ bill_length_mm + flipper_length_mm + species,
  data = p_train
)

# model with interaction between bill_length_mm and species
model_int <- linear_reg() %>% set_engine("lm") %>% fit(
  body_mass_g ~ bill_length_mm * species + flipper_length_mm,
  data = p_train
)

# evaluate on train set
pred_no_int_train <- predict(model_no_int, p_train) %>% bind_cols(p_train)
pred_int_train    <- predict(model_int, p_train) %>% bind_cols(p_train)

metrics_no_int_train <- pred_no_int_train %>% reg_metrics(truth = body_mass_g, estimate = .pred)
metrics_int_train    <- pred_int_train %>% reg_metrics(truth = body_mass_g, estimate = .pred)

# evaluate on test set
pred_no_int_test <- predict(model_no_int, p_test) %>% bind_cols(p_test)
pred_int_test    <- predict(model_int, p_test) %>% bind_cols(p_test)

metrics_no_int_test <- pred_no_int_test %>% reg_metrics(truth = body_mass_g, estimate = .pred)
metrics_int_test    <- pred_int_test %>% reg_metrics(truth = body_mass_g, estimate = .pred)

# print results
cat("=== No interaction: Training ===\n"); print(metrics_no_int_train)
cat("\n=== Interaction: Training ===\n"); print(metrics_int_train)
cat("\n=== No interaction: Test ===\n"); print(metrics_no_int_test)
cat("\n=== Interaction: Test ===\n"); print(metrics_int_test)
```

Finally, let's visualize

```{r}
# visualization: test set predicted vs observed for both models
p1 <- ggplot(pred_no_int_test, aes(x = body_mass_g, y = .pred, color = species)) +
  geom_point(alpha = 0.8) + geom_abline(lty = 2) +
  labs(title = "No Interaction: Test Set", x = "Observed", y = "Predicted")

p2 <- ggplot(pred_int_test, aes(x = body_mass_g, y = .pred, color = species)) +
  geom_point(alpha = 0.8) + geom_abline(lty = 2) +
  labs(title = "With Interaction: Test Set", x = "Observed", y = "Predicted")

# Print side-by-side (requires patchwork or cowplot). If not available, print separately.
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p1 + p2)
} else {
  print(p1); print(p2)
}

# Optional: inspect interaction coefficients
tidy(model_int)

```

Generally: 

- if the interaction model lowers test RMSE, the interaction helped generalization. here, the interaction term allows the slope between bill length and body mass to differ by species — biologically reasonable if species have different scaling relationships.
- also it's worth checking whether the interaction model greatly improves training fit but not test fit (overfitting warning).

:::
